---
title: "Session 02. Data Management"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Data Structures


## Atomic vectors: Creation, Dimensionality, Homogeneity

```{r}
logic <- c(TRUE, FALSE, FALSE)
typeof(logic)
```

```{r}
int <- c(-1L, 99L)
typeof(int)
```

```{r}
# Explain the result
dbl <- sqrt(8L)
typeof(dbl)
```

```{r}
# Explain the result
chr <- c(TRUE, "Hello", 2.0, "R")
typeof(chr)
```


## Atomic Vectors: Attributes

```{r}
die <- 1:6; length(die) # Check the length of an atomic vector
```

```{r}
length(c(die, die)) # Will the length differ?
```

```{r}
names(die) # Explain the output.
```

```{r}
names(die) <- c("one", "two", "three", "four", "five", "six")
die
```

```{r}
die + 1  # Explain the result
```

```{r}
names(die) <- NULL # Reset the attribute
die
```


## Data Frames: Creation

- **Creation**
```{r}
students <- data.frame(
  first_name = c("Alex", "Jessy", "Barbara", "Jacob"),
  student_id = c(349857, 796245, 143577, 987456),
  passed = c(TRUE, TRUE, FALSE, FALSE),
  stringsAsFactors = FALSE # Will be explained later
)
students
```


## Data Frames: Dimensionality, Homogeneity, Attributes

Use str() to get a brief description of the data frame. By the way, RStudio reports the same information in the Environment pane.

```{r}
str(students)
```

The following code block demonstrates how you query the dimensions of a data frame.

```{r}
dim(students); nrow(students); ncol(students)
```

Here is how you query its name atributes.

```{r}
rownames(students); names(students) # also: colnames()
```


# R Notation

In the following, we are going to explore *R*'s different indexing systems. Let's load some toy data first. Later, we are going to hone our data management skills on these data.

```{r}
student_data <- read.csv("./dta/student_data.csv", stringsAsFactors = FALSE)
# read.csv() loads **comma**-separated flat files. By
# default, Excel returns semi-colon separated csv files.
# Use read.csv2 to read those into memory.
head(student_data)
```


## Integers


### Positive Integers

Select the first row and column.
```{r}
student_data[ , ]
```

Select rows 1 through 5 & colums 1 through 3.

```{r}
student_data[ , ]
```

Try to explain the result. Could that feature be useful?
```{r}
student_data[0, 0]
```


### Negative Integers

Deselect the first row and column.
```{r}
student_data[ , ]
```

Select rows 1 through 5 and drop columns 2 to 5.
```{r }
student_data[ , ] 
```

This statement returns an error. Explain why.
```{r }
student_data[-3:4, 1:2]
```


## Blanks

Select the first row and all columns.

```{r }
student_data[ , ]
```

Select all rows from the second and fifth column.

```{r }
student_data[ , ] 
```

Return the entire data frame

```{r }
student_data[ , ]
```


## Logical Indexing

Explain the result.

```{r}
student_data[FALSE , ]
```

Only select columns 1, 2, 6

```{r}
mask <- c( , , , , , )
student_data[ , mask]
```

Execute the statement and explain the result.

```{r}
student_data[student_data[, 2] == 1, ]
```


## Names

Select columns student and student_id.

```{r}
student_data[ , ]
```

Execute the statement and explain the error.

```{r }
student_data[ , "Student"] 
```

Execute the statement and explain the error. Can you come up with a working solution?

```{r}
student_data[ , -"student_id"]
```


## Dollar Signs & Double Brackets

Return column arts.

```{r}
student_data$
```

Return the first five student ids.

```{r}
student_data$
```

Extract column gender.

```{r}
student_data[[ ]]
```

The following statement returns an error. Try to explain why.

```{r}
student_data[[c("student", "student_id")]]
```
