---
title: "Session 02: Data Management Challenge"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

# Load The Data

```{r}
student_data <- read.csv("./dta/student_data.csv",
    stringsAsFactors = FALSE)
```

# Part A

Some grades are missing. Angela scored 603 in Science,
and Cheryl 28 in Literature. Find and fill the gaps.

```{r}
student_data[ , ] <- 603
student_data[ , ] <- 28
```

Joel withdrew from class. Remove him from the dataset.

```{r}
student_data <- student_data[ , ]
```

Gender was coded accidentally. Drop the variable.

```{r}
student_data <- student_data[ , ]
```

The grades are on widely different scales. **Scale** them to comparable units.

```{r}
tmp <- student_data[, c("science", "arts", "literature")]
tmp <- ...(tmp)
```

# Part B

Take the average of scaled science, arts, and literature grades for each student.

```{r}
grade <- ...
student_data <- data.frame(student_data, grade = grade)
rm(tmp, grade)
```

Find its .2, .4, .6, and .8 quantiles. Assign alphabetical grades.

```{r}
grade_quantiles <- ...(student_data[, "grade"], ... = c(.2,.4,.6,.8))
# Hint:
# A: grade > .8
# B: .8 <= grade & grade > .6
# C: .6 <= grade & grade > .4
# D: .4 <= grade & grade > .2
# F: grade < .2
student_data[..., "grade_alp"] <- "F"
student_data[..., "grade_alp"] <- "D"
student_data[..., "grade_alp"] <- "C"
student_data[..., "grade_alp"] <- "B"
student_data[..., "grade_alp"] <- "A"
```

Greg was caught cheating. He should receive an F.

```{r}
student_data[..., "grade_alp"] <- ...
```

Create a new data frame grade_data. This data frame should:

- include only student_id and the final grade;
- be ordered by student_id.

```{r}
office_list <- student_data[ , ]
office_list <- student_data[...(student_data[["student_id"]]), ]
# ATTENTION: Do not fall for sort()!
office_list
```

Congratulations! That wasn't exactly easy.