---
title: "Session 03: Data Management Challenge"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

# Sequential Execution

A script runs step by step from start to end (or any other condition is met).

```{r}
A <- "Well"
B <- "hello there."
paste(A, B, sep = ", ")
rm(A, B, C) # Explain the error.
```


# Iterative Execution


## for()-loops

Execute statement(s) repeatedly over a set of values.

```{r}
for (value in c("Waiting", "for", "statistics.")) {
  print(value)
}
```

It is good practice to loop over a set of integers which indexes both object and storage simultaneously.

```{r, eval = FALSE}
words <- c("So", "how's", "looping", "so", "far?")
chr <- vector("character", length = length(words))
chr
for (i in 1:length(words)){
    cat("Iteration variable i is now:", i, "\n")
    chr[i] <- words[i]
}
chr
```


## Exercise

Create a for()-loop to replace your assign letter grades with a more concise statement. Use the quantile-based assignment rule below.

Quantile | >=0 | >=.2 | >=.4 | >=.6 | >=.8
Grade    |   F |    D |    C |    B |    A

Answer the following questions first:

1. What should we loop over?
2. Where should we store results?

```{r, eval = FALSE}
set.seed(12345678)
grades <- rnorm(38, 0, 1)  # 38 grades ~ N(0, 1)
letter_grades <- vector("character", length(grades))
Letters <- c("F", "D", "C", "B", "A")
# non-standard notation -> try letters and LETTERS in the console
```


# Conditional Execution

## if()-Statements

A simple absolute value algorithm:

```{r}
x <- -4
if (x < 0) {
  x <- -1 * x
}
x
```

What will these statements return? Think first, check later.

```{r}
x <- 1
if (TRUE) {
  x <- 2
}
x
```

```{r}
x <- 1
if (x == 1) {
  x <- 2
  if (x == 1) {
    x <- 3
  }
}
```


## else()-Statements

Let's round π.

```{r}
dec <- pi
if ((dec - trunc(dec)) >= 0.5) {
  dec <- trunc(dec) + 1
} else {
  dec <- trunc(dec)
}
dec
```


# Practical Challenge

Estimate π from the chance to hit a circle perfectly inscribed in a square with a randomly thrown dart. Your estimate will be $4 * p(Hit the circle)$. Here are some hints.

- A circle is a shape consisting of **all points in a plane** that are **a given distance from a given point**. So, where is your origin and what's your radius?
- By default, runif() samples from the unit interval [0, 1], but the unit is not necessarily what you want.
- Did your dart land inside the circle? Use the Euclidian vector norm $\sqrt{\sum_{i=1}^{N}x_i^2}$ to find out.
- The precision of your estimate increases with the number of darts
you throw. Don't overshoot it though.
- You know enough *R* to solve this problem in four lines of code!
