---
title: "Session 5. Basic Statistics"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

```{r, setup}
packages <- c("tidyverse")
for (package in packages){
    if (!require(package, character.only = TRUE)) {
        install.packages(package, repos = "https://cloud.r-project.org")
        library(package, character.only = TRUE)
    }
}
rm(package, packages)
asoiaf <- read.csv2("./dta/asoiaf.csv", stringsAsFactors = FALSE)
head(asoiaf)
```

# Many Values

## Central Tendency and Spread

Execute each of the following commands. Explain the output. Set na.rm to FALSE. What happens?

```{r, eval = FALSE}
mean(asoiaf[, "age_in_chapters"], na.rm = TRUE)
```

```{r}
sd(asoiaf[, "age_in_chapters"], na.rm = TRUE)
```

```{r}
quantile(asoiaf[, "age_in_chapters"],
  probs = c(0, .01, .05, .25, .5, .75, .9, .95, 1),
  na.rm = TRUE
)
```

```{r}
summary(asoiaf[, "age_in_chapters"])
```

## Graphical Exploratory Data Analysis

ggplot2 offers numerous exploratory graphs.[^2] Create
each of the graphs below. What do they return?

```{r, eval = FALSE}
p <- ggplot(data = asoiaf, aes(x = age_in_chapters))
```

```{r}
p + geom_histogram()
```

```{r}
p + geom_density() + labs(y = "PDF")
```

```{r}
p + stat_ecdf() + labs(y = "CDF")
```

```{r}
p + geom_boxplot(aes(x = 0, y = age_in_chapters))
```

```{r}
ggplot(data = asoiaf, aes(sample = age_in_chapters)) +
  geom_qq() + geom_qq_line()
```

## Grouping Values

We regularly decrease the granularity of many-valued variables in some controlled fashion. Frequently, we calculate grouped statistics to that end. On other occasions, we discretize continous variables. The following chunks demonstrate how to do either.

Use aggregate() to return single value statistics by group.

```{r}
mu_age_by_allegiance <- aggregate(
  x = asoiaf[, "age_in_chapters"],
  by = list(allegiances = asoiaf[, "allegiances"]),
  FUN = mean, na.rm = TRUE
)
mu_age_by_allegiance
```

Use cut(), or even better: Hmisc::cut2() to discretize continuous variables in different ways.

```{r}
tmp <- cut(x = asoiaf[, "age_in_chapters"], breaks = 5)
# generate <breaks> pieces of equal length
summary(tmp) # Do you notice anything weird?
```

The number of observations differs across categories. We can supply
quantile values to avoid such skewed discretization of numerical data.

```{r}
tmp <- cut(x = asoiaf[, "age_in_chapters"],
  breaks = quantile(
    asoiaf[, "age_in_chapters"], na.rm = TRUE
  ), # vector of values at which to cut x.
  include.lowest = FALSE
)
summary(tmp) # Notice something weird?
```


# Few Values

## What are factors?

```{r, results=FALSE}
grades <- c("A", "B", "B", "C")
grades <- factor(grades,
  levels = c("C", "B", "A"), # state ALL values
  labels = c("C", "B", "A"), # name EACH value
  ordered = TRUE # defaults to FALSE (nominal data)
)
grades
```

```{r}
typeof(grades)
```

```{r}
attributes(grades)
```

Factors are a neat and memory-efficient way to store categorical data. Be careful
when working with factors, though because they can behave in unexpected ways.

Under the hood, integers represent individual factor levels.
```{r}
as.numeric(grades)
```

However, we use their string labels to operate on factors.
```{r}
grades > "C" # will work fine
```

```{r}
grades > 1 # will generally not work
```


## Simple N-way Contingency Tables

```{r}
table(asoiaf[, "gender"]) # single 1way table
```

```{r}
apply( # Multiple 1way tables
  asoiaf[, c("gender", "nobility")], 2, table
)
```

```{r}
table( # Create a Bivariate Table
    "Book of Death" = asoiaf[, "book_of_death"], # Naming is optional, but helpful
    "Nobility" = asoiaf[, "nobility"]
)
```

```{r}
mytable <- table(
  "gender" = asoiaf[, 'gender'],
  "nobility" = asoiaf[, 'nobility']
)
prop.table(mytable)
# Add argument margin = {1; 2}. What happens?
addmargins(mytable)
# Add argument margin = {1; 2}. What happens?
```


# Adventures in Association

```{r}
mytable <- table(
  "gender" = asoiaf[, 'gender'],
  "nobility" = asoiaf[, 'nobility']
)
```

```{r}
fit <- chisq.test(mytable); fit # Chi-Square Test
```

```{r}
fit <- fisher.test(mytable); fit # Fisher's Exact Test
```


## Correlation Analysis

```{r}
# a. Explain the code and plot. ---------------------
ggplot(data = asoiaf,
  aes(x = chapter_of_intro, y = chapter_of_death)
) + geom_point() +
  geom_smooth(aes(col = "loess"), method = "loess") +
  geom_smooth(aes(col = "ols"), method = "lm")
```

```{r}
# b. Correlation Analysis ---------------------------
cor(x = asoiaf[, c(6, 8)],
  use = "complete.obs", # What does <use> do?
  method = "pearson" # {pearson; kendall; spearman}
) # Now try cor.test() on your own.
```

## Mean Comparison Tests

- Question: Do two groups come from the same population?
- In ASOIAF: Do nobles survive longer than other social strata?

```{r, eval = FALSE}
t.test( # Alternative: mu_0 < mu_1
  age_in_chapters ~ nobility, data = asoiaf,
  alternative = "less" # "greater" for mu_0 > mu_1
)
```
