---
output:
  beamer_presentation:
    theme: Berlin
    colortheme: seagull
    fonttheme: professionalfonts
    highlight: tango
    toc: false
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("ggplot2")
asoiaf <- read.csv2(
  "~/github/teaching/pcqr/r_intro/05/dta/asoiaf.csv",
  stringsAsFactors=FALSE
)
```

\maketitle

# Introduction

## So far, you have learned...

- to navigate the $\mathcal{R}$ environment,
- to create, index, and modify objects,
- to take advantage of control flow statements,
- generate decent looking graphs.

Looks like you are ready to unleash $\mathcal{R}$ on real data.

## What are we going to do?

We are going to plunge head first into data analysis, taking
advantage of data on all *named* characters in George R.R. Martin's
"A Song of Ice and Fire".[^1] In the end, you will be able to offer
scientific insight on the erstwhile mother of all cocktail party
questions: **What's the chance that Jon Snow is going to die?**

To get started:

1. Quit & reopen $\mathcal{R}$.
2. Using RStudio open "./05/notebook_05.Rmd"
3. Execute the first code chunk.

[^1]: O'Neill, M. 2016. Game of Thrones. https://bit.ly/2qjUfQ2 (last access: 10/08/2018).

## Outline
\tableofcontents

## Few or Many Values?

We are interested in the variables allegiances,
age_in_chapters, gender, and nobility.

- Proper tools for description change with data type
- Indicator: How many different values does a variable take?

# Many Values

## Central Tendency and Spread

Execute each of the following commands. Explain the output. Set na.rm to FALSE. What happens?

```{r, eval = FALSE}
mean(asoiaf[, "age_in_chapters"], na.rm = TRUE)
sd(asoiaf[, "age_in_chapters"], na.rm = TRUE)
quantile(asoiaf[, "age_in_chapters"],
  probs = c(0, .01, .05, .25, .5, .75, .9, .95, 1),
  na.rm = TRUE
)
summary(asoiaf[, "age_in_chapters"])
```

>- **What have we learned?**
>- mean(), sd(), and quantile() return just that.
>- Each requires instructions on how to process NAs.
>- summary() returns the 5-point-summary plus mean and NAs.

## Graphical EDA

ggplot2 offers numerous exploratory graphs.[^2] Create
each of the graphs below. What do they return?

```{r, eval = FALSE}
p <- ggplot(data = asoiaf, aes(x = age_in_chapters))
p + geom_histogram()
p + geom_density() + labs(y = "PDF")
p + stat_ecdf() + labs(y = "CDF")
p + geom_boxplot(aes(x = 0, y = age_in_chapters))
ggplot(data = asoiaf, aes(sample = age_in_chapters)) +
  geom_qq() + geom_qq_line()
```

[^2]: For an entire theory of graphical EDA using
  ggplot2 see Unwin, A. 2015. Graphical Data Analysis
  with R. Boca Raton: CRC Press.

## Grouping Values

- **Goal**: Controlled loss of information for, e.g., tables
- **Means**:
    - Aggregate by some meaningful group
    - Discretize continuous variables

```{r, eval = FALSE}
# Variant a. Aggregate data --------------------------
mu_age_by_allegiance <- aggregate(
  x = asoiaf[, "age_in_chapters"],
  by = list(allegiances = asoiaf[, "allegiances"]),
  FUN = mean, na.rm = TRUE
); mu_age_by_allegiance
```

## Grouping Values, contd.

```{r, eval = FALSE}
# Variant b. Discretize the data -------------------------
tmp <- cut(x = asoiaf[, "age_in_chapters"],
  breaks = 5
  # divides data into <breaks> pieces of equal length
); summary(tmp) # Note something weird?
tmp <- cut(x = asoiaf[, "age_in_chapters"],
  breaks = quantile(
    asoiaf[, "age_in_chapters"], na.rm = TRUE
  ), # vector of values at which to cut x.
  include.lowest = FALSE
); summary(tmp) # Notice something weird?
typeof(tmp); class(tmp) # Notice something weird?
```

# Few Values

## What are factors?

- Special instance ("class") of atomic vectors
- Store nominal and ordinal data, e.g., eye color & letter grades
- Look like character strings, but may behave like integers
- Create factors only when needed

```{r, results=FALSE}
grades <- c("A", "B", "B", "C")
grades <- factor(grades,
  levels = c("C", "B", "A"), # state ALL values
  labels = c("C", "B", "A"), # name EACH value
  ordered = TRUE # defaults to FALSE (nominal data)
)
typeof(grades); attributes(grades) # Try these.
```

## What are factors?, contd.

- Components of a factor: numeric value & character label
- labels (BUT NOT VALUES) can be used for logical indexing

```{r}
grades; as.numeric(grades)
grades > "C" # will work fine
grades > 1 # will generally not work
```

## Simple N-way Contingency Tables

- table() creates N-way contingency tables

```{r}
table(asoiaf[, "gender"]) # single 1way table
```


Explain the output of these statements. Explain the difference in
your own words.

```{r, eval = FALSE}
apply(
  asoiaf[, c("gender", "nobility")], 2, table
)
table(asoiaf[, "book_of_death"], asoiaf[, "nobility"])
```

## Add Information to Contingency Tables

- You must state explicitly what information you require.
- Examples: Proportions & Totals

```{r, eval = FALSE}
mytable <- table(
  "gender" = asoiaf[, 'gender'],
  "nobility" = asoiaf[, 'nobility']
)
prop.table(mytable)
# Add argument margin = {1; 2}. What happens?
addmargins(mytable)
# Add argument margin = {1; 2}. What happens?
```

# Adventures in Association

## Two-way Contingency Tables

- Numerous methods provided
- Most defined by individual functions
- See packages [vcd & vcdExtra](https://cran.r-project.org/web/packages/vcdExtra/vignettes/vcd-tutorial.pdf) for more options

```{r, results = FALSE}
mytable <- table(
  "gender" = asoiaf[, 'gender'],
  "nobility" = asoiaf[, 'nobility']
)
fit <- chisq.test(mytable); fit # Chi-Square Test
fit <- fisher.test(mytable); fit # Fisher's Exact Test
```

## Correlation Analysis

- Scatter plots are the starting point for any correlation analysis
- base::pairs() & car::scatterplotMatrix() return plot matrices
- Quantify associations using cor() and cor.test() functions

```{r, eval = FALSE}
# a. Explain the code and plot. ---------------------
ggplot(data = asoiaf,
  aes(x = chapter_of_intro, y = chapter_of_death)
) + geom_point() +
  geom_smooth(aes(col = "loess"), method = "loess") +
  geom_smooth(aes(col = "ols"), method = "lm")
# b. Correlation Analysis ---------------------------
cor(x = asoiaf[, c(6, 8)],
  use = "complete.obs", # What does <use> do?
  method = "pearson" # {pearson; kendall; spearman}
) # Now try cor.test() on your own.
```

## Mean Comparison Tests

- Question: Do two groups come from the same population?
- In ASOIAF: Do nobles survive longer than other social strata?

```{r, eval = FALSE}
t.test( # Alternative: mu_0 < mu_1
  age_in_chapters ~ nobility, data = asoiaf,
  alternative = "less" # "greater" for mu_0 > mu_1
)
```

- Defaults to unequal variances & returns Welch's t-test
- Set paired = TRUE for paired t-tests
- For detailed instructions see Field, A. et al. 2012.
  Discovering Statistics Using R. Los Angeles: SAGE, Ch. 9.

# Summary

## R functions for basic statistics\footnote{
  Selectively adapted from Fox, J. and S. Weisberg. 2011.
  An R Companion to Applied Regression. 2nd ed. London: SAGE.
}

\resizebox{\textwidth}{!}{
  \centering
  \begin{tabular}{*{2}{l}}
  Method & $\mathcal{R}$ Function(s) in base\\
  \midrule
  Mean & mean()\\
  Median & median()\\
  Quantiles & quantile()\\
  Extremes & range(), min(), max()\\
  Variance & var()\\
  Standard Deviation & sd()\\
  Covariance Matrix & vcov()\\
  Correlation & cor() \\
  Analysis of Variance & aov(), lm(), anova()\\
  Contingency Tables & table(), xtabs() \\
  t-test for Means & prop.test(), binom.test() \\
  $\chi^2$ Test for Indep. & chisq.test() \\
  Various Nonparametric Tests & friedman.test(), kruskal.test(), wilcox.test(), etc.
  \end{tabular}
}

