---
title: "Session 6. (G)LMs"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

```{r}
packages <- c("tidyverse")
for (package in packages){
    if (!require(package, character.only = TRUE)) {
        install.packages(package, repos = "https://cloud.r-project.org")
        library(package, character.only = TRUE)
    }
}
rm(package, packages)
asoiaf <- ...("./06/dta/asoiaf.csv", ...)
head(asoiaf)
asoiaf[, "is_dead"] <- !is.na(asoiaf[, "book_of_death"])
```

**Exercise** How would you write the following formulas?[^1]

1. $y = a + x + z + xz$
2. $y = a + x + x^2 + x^3$
3. $log_{10}(y) = x + z + w + xz + xw + wz$
4. $y$ as a function of variables in the data but $k$

[^1]: Assume $a$ is the constant.


## Get Your Hands Dirty

Now it's your turn. Use the **asoiaf** data to

- regress **is_dead** on
- **allegiances**,
- the full interaction of **gender** and **nobility**, and
- a cubic polynomial on **age_in_chapters**.
- Save the results to an object called **myfit**.

```{r}
myfit <- ...(
  formula = ...,
  data = asoiaf
)
```

# Working With Results

## Generic Sequence

1. Define scenarios to predict
2. Create a date frame which contains those scenarios
3. Use predict() to return quantities of interest
4. Summarize the results


## Example: How does the chance to die change over time?

```{r}
# Steps 1 & 2 Prediction scenarios
pred_dta <- data.frame(
  allegiances = "Baratheon",
  gender = mean(asoiaf$gender),
  nobility = mean(asoiaf$nobility),
  age_in_chapters = 1:343, stringsAsFactors = FALSE
)
head(pred_dta)
```

```{r}
# Step 3 Predict
pred_dta[, "fitted"] <- predict(myfit, newdata = pred_dta)
head(pred_dta)
```

```{r}
# Step 4 Summarize
ggplot(data = pred_dta,
  aes(x = age_in_chapters, y = fitted)) + geom_line()
```


# Generalized Linear Models

```{r}
myfit <- ...(
  formula = is_dead ~ 0 + allegiances +
    gender * nobility +
    age_in_chapters + I(age_in_chapters^2) +
    I(age_in_chapters^3),
  family = ...,
  data = asoiaf
)
```

## Is Jon Snow going to die?

```{r}
jon_snow <- which(asoiaf$name == "Jon Snow")
pred_dta <- asoiaf[rep(jon_snow, 3), ]; rm(jon_snow)
pred_dta[2, "allegiances"] <- "Stark"
pred_dta[3, "allegiances"] <- "Targaryen"
pred_dta[, "fitted"] <- predict(
  myfit, newdata = pred_dta, type = "response"
)
pred_dta[, "fitted"]
```

# Regression Diagnostics

```{r, eval = FALSE}
plot(myfit)
```