---
title: "Session 01: The Very Basics"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# The Very Basics


## Let's take *R* for a ride.

Use *R* to answer the following problems. Remember:  **Strg+Enter** executes a chunk line, **Strg+Shift+Enter**
executes the entire chunk. Your output will appear immediately below the chunk.

```{r}
# Add 2 and 2.

```

```{r}
# Multiply -3 and 9.

```

```{r}
# Find the cubic root of 8.

```

```{r}
# Find the natural log of 1.

```

Finally, *R* ships with an extensive online documentation. For instance, it includes a manual named ``An Introduction to R''. Use the RStudio help pane and open that manual.


## What is an object?

*R* stores data in objects. When encountering an object, *R* returns the data saved inside. Think of objects as the nouns of the *R* language.

```{r}
a <- 1 # Assign value "1" to object "a".
a # Retrieve the value of "a".
```

```{r}
a + 2 # Retrieve the value of "a" and add "2" to it.
```

```{r}
a <- 999; a # On reassignment R overwrites an object.
```


## The Rules of Assignment

There are several valid assignement patterns.

```{r}
a <- 3 # Object <- Value
a
```

```{r}
9 -> a # Value -> Object
a
```


## Naming Ojects

- Rule 1: An object name cannot start with a number.
- Rule 2: An object name cannot use certain special symbols,
      e.g., ^, !, $, @, +, -, /, [, or reserved names, e.g., NA, NULL, etc.
- Rule 3: Capitalization matters, i.e. A ≠ a.

```{r }
!d0 <- 1
```

```{r}
FOO <- "bar"
```

```{r }
1_day <- 1
```

```{r }
day_1 <- TRUE
```

```{r }
_day1 <- FALSE
```

```{r }
day.1 <- "99"
```

```{r }
day^1 <- 0
```

## How does *R* are operate on objects?

Execute these commands and describe the result.

```{r}
die <- 1:6; die * die
```

```{r}
die + 1:3
```

```{r}
die %*% die
```

## What is a function?

*R* uses functions to operate on data. Think of functions as the verbs of the *R* language. The data passed to a function is called its argument. Execute the following commands and describe the result.

```{r, eval = FALSE}
die <- 1:6; mean(die)
```

```{r}
round(mean(1:6), digits = 2) 
```

```{r}
round(mean(die), 2)
```

